---
title: "Drug Abuse"
author: "Oscar Briones Ramirez"
date: '2023-04-09'
output: html_document
---

```{r setup, include=FALSE}
setwd("/Users/oscar/Desktop/Winter2023/STAT469/DrugAbuse")

library(tidyverse)
library(GGally)
library(car)
library(MASS)
library(lmtest)
library(multcomp)
library(forecast)
library(gstat)
library(nlme)
library(sf)
library(spdep)

source("moranBasis.R")
```

## Reading in and Handling Shapefiles

**1. Read in the drug abuse shapefile.**
```{r readshp, include=TRUE}
myShp <- st_read("DrugAbuse.shp")
```


**2. Create a dataframe from the shapefile that exclude the polygon information.**
```{r datafr, include=TRUE}
myShpDF <- data.frame(myShp)%>%
  dplyr::select(-geometry)
```


## Exploratory Data Analysis

**1.Create a pairs plot (ggpairs()) to assess the relationship between log(Count) and the explanatory variables (note we are using log(Count) here as the response because Poisson regression is log-linear).**

```{r ggpairs, include=TRUE}
ggpairs(log(myShpDF[,-1]))
```

**2. Create a choropleth map of log(Count).**
```{r chlor, include=TRUE}
ggplot(data=myShp) +  geom_sf(mapping=aes(fill=log(Count)), color="black") +  scale_fill_distiller(palette="Spectral")
```

**3.Fit a lm() of log(Count) using Population, PctWhite, MedInc and Pct18to34 as explanatory variables. Perform a Moran’s I test on the residuals to see if there is spatial correlation in the residuals.**
```{r lm, include=TRUE}
drug.lm <- lm(formula=log(Count)~Population + PctWhite + MedInc + Pct18to34, data=myShpDF)

moran.test(x=drug.lm$residuals, listw=nb2listw(poly2nb(myShp)))
```
**4. Perform a Geary’s C test on your residuals from #3 above to double check if there is spatial correlation in the residuals.**

```{r geary, include=TRUE}
geary.test(x=drug.lm$residuals, listw=nb2listw(poly2nb(myShp)))
```
**5. Map the residuals from the lm() fit to see if there is spatial correlation.**
```{r map1, include=TRUE}
ggplot(data=myShp) +
  geom_sf(mapping=aes(fill=resid(drug.lm)))
```

## Defining Spatial Basis Functions

**1. Create the adjacency matrix.**
```{r adjmat, include=TRUE}
A <- nb2mat(poly2nb(myShp), style="B")
```

**2. Create the Moran spatial basis and plot the first basis in a chloropleth map.**
```{r moran1, include=TRUE}
X <- model.matrix(object=log(Count)~Population + PctWhite + MedInc + Pct18to34, data=myShpDF)

M <- moranBasis(X, A, tol=0.95)

ggplot(data=myShp) +
  geom_sf(mapping=aes(fill=M$B1))
```

**3. Merge the Moran spatial bases into your myShpDF data frame for use in fitting models later.  **

```{r cbind, include=TRUE}
myShpDF <- bind_cols(myShpDF, M)
```


## Spatial GLM Model Fitting

**1. Fit a spatial GLM model with Count as the response and using PctWhite, MedInc, Population, Pct18to34 AND your spatial bases as explanatory variables. Print a summary() of the model to see your coefficient table.**

```{r glm, include=TRUE}
drug.glm <- glm(formula=Count~. - State, data=myShpDF, family=poisson)
summary(drug.glm)
```

## Validating Spatial MLR Model Assumptions and Predictions

**1.Check the assumption of linearity using added-variable plots.**

```{r avplots, include=TRUE}
avPlots(drug.glm, ask=FALSE)
```

**2. Check the assumption of independence by decorrelating residuals and performing Moran’s I or Geary’s C tests to make sure there is no more spatial correlation.**
```{r I, echo = TRUE}
source("stdres.gls.R")
sres <- stdres.gls(drug.glm)

moran.test(x=sres, listw=nb2listw(poly2nb(myShp)))
```
**3.Draw a choropleth map of the standardized and decorrelated residuals to visually verify that the residuals are no longer spatially correlated.**

```{r clordecor, echo = TRUE}
ggplot(data=myShp) +
   geom_sf(mapping=aes(fill=sres), color="black") +  scale_fill_distiller(palette="RdBu")
```

**4. Check the assumption of equal variance by plotting the standardized and decorrelated residuals vs. the log(fitted values).**

```{r eqvar, echo = TRUE}
ggplot(mapping=aes(y=log(fitted(drug.glm)), x=sres)) + geom_point()
```

## Statistical Inference

**1. Print out the summary of the GLM model fit and identify the estimates and 95% confidence intervals of your explanatory variables. **
```{r summ, echo = TRUE}
summary(drug.glm)

confint(drug.glm,level=.95)
```

**2. Create a chloropleth map of the spatially correlated residuals (just the b′iθ̂ part) to identify states that, after accounting for the explanatory variables, have an elevated level of risk.**

```{r chlorcor, echo = TRUE}

spatial <- as.matrix(coef(drug.glm)[-(1:5)])

M <- as.matrix(M)


ggplot(data=myShp) +
   geom_sf(mapping=aes(fill=M %*% spatial), color="black") +  scale_fill_distiller(palette="RdBu")


```